Practical Optional Types for Clojure

### 2015-07-05 intro.tex
More editing.
New page for bib. Changes to sec 1 & 2

\begin{figure}[t!]
\inputminted[firstline=5,lastline=12]{clojure}{code/demo/src/demo/ex-intro.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

@@

A simple example of these features in combination is given in
figure~\ref{fig:ex1}. Here, the \texttt{tostr} multimethod dispatches
on the \texttt{class} of the argument---when it is a \texttt{String},
the first method implementation is called, for a Java \texttt{File}, the
second. The method for \texttt{File} uses Java interoperation to call
the \texttt{.getPath} method, and Typed Clojure uses the type
information to help determine which JVM method call to construct.

Finally, when not working with existing Java classes and objects, the
most common Clojure data structure is the dictionary, a
high-performance immutable table which Clojure programmers repurpose
for all manner of data types. Simply treating them as uniformly-typed
key-value mappings would be entirely insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous maps}, in which mandatory, optional, and
absent keys can be specified.

@@

\begin{itemize}
  \item We motivate and describe  Typed Clojure, an optional
    type system for Clojure that understands existing Clojure idioms.
  \item We present a sound formal model for the three crucial type
    system features that Typed Clojure relies on: multi-methods, Java
    interoperability, and heterogenous maps.
  \item We evaluate the use of Typed Clojure features on existing
    Typed Clojure code, including both open source and in-house systems.
\end{itemize}

@@

 The remainder of this papers begins with an example-driven
 presentation of the main type system features in
 \secref{sec:overview}. We then incrementally present a core calculus
 for Typed Clojure covering all of these features together in
 \secref{sec:formal} and prove type soundness
 (\secref{sec:metatheory}). We then discuss the full implementation of
 Typed Clojure, dubbed \coretyped{}, which extends the formal model in
 many ways, and empirical analysis of significant code bases written
 in \coretyped{} in \secref{sec:experience}. Finally, we discuss
 related work and conclude.

>>>>

\begin{figure}[t!]
\inputminted[firstline=5]{clojure}{code/demo/src/demo/ex-intro.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

@@

A simple example of these features in combination is given in
figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
on the \texttt{class} of the argument---when it is a \texttt{String},
the first method implementation is called, for a Java \texttt{File}, the
second. The method for \texttt{String} uses Java interoperation
to call a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
method for \texttt{File} invokes
the \texttt{.getName} method with a non-nil target, returning a nilable
type. Typed Clojure uses type information to resolve JVM method or constructor overloads,
avoiding expensive runtime reflective calls.

Finally, when not working with existing Java classes and objects, the
most common Clojure data structure is the dictionary, a
high-performance immutable table which Clojure programmers repurpose
for all manner of data types. Simply treating them as uniformly-typed
key-value mappings would be entirely insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous map} type, in which mandatory, optional, and
absent keys can be specified.

@@

\begin{enumerate}
  \item We motivate and describe  Typed Clojure, an optional
    type system for Clojure that understands existing Clojure idioms.
  \item We present a sound formal model for the three crucial type
    system features that Typed Clojure relies on: multi-methods, Java
    interoperability, and heterogenous maps.
  \item We evaluate the use of Typed Clojure features on existing
    Typed Clojure code, including both open source and in-house systems.
\end{enumerate}

@@

\noindent
 The remainder of this paper begins with an example-driven
 presentation of the main type system features in
 \secref{sec:overview}. We then incrementally present a core calculus
 for Typed Clojure covering all of these features together in
 \secref{sec:formal} and prove type soundness
 (\secref{sec:metatheory}). We then discuss the full implementation of
 Typed Clojure, dubbed \coretyped{}, which extends the formal model in
 many ways, and empirical analysis of significant code bases written
 in \coretyped{} in \secref{sec:experience}. Finally, we discuss
 related work and conclude.

### 2015-07-05 overview.tex
New page for bib. Changes to sec 1 & 2

\subsection{Typed Clojure Introduction}

@@

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n]
  (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
(greet nil)    ; Type Error:
               ;   Expected Str, given nil
\end{minted}

The function may take strings, but providing \clj{nil} (Clojure's name for Java's \emph{null})
is a static type error --- \clj{nil} is not a string in Typed Clojure.

@@

All Clojure values are true, except \clj{nil} and \clj{false}, so the
comma is only added when the argument is non-nil.

\paragraph{Flow analysis} Typed Clojure uses occurrence typing to
model type-based control flow.

@@

Providing \clj{repeat} with a nilable first argument is disallowed,
but the \clj{when} expression ensures it can only be passed an integer.

@@

@@

Clojure supports interoperability with Java, including the ability to
call constructors, methods and access fields.

\begin{minted}{clj}
  (fn [f] (.getParent f))
\end{minted}

Calls to Java methods and fields have prefix notation
like \clj{(.method target args*)} and \clj{(.field target)} respectively,
with method and field names prefixed with a dot and methods taking some number of arguments.

\begin{exmp}
\inputminted[firstline=18,lastline=19]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

Typed Clojure and Java treat \java{null} differently.
In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
This means the Java static type \java{String} is equivalent to
\clj{(U nil String)} in Typed Clojure.

Reference types in Java are nullable, so to guarantee a method call does not
leak \java{null} into a Typed Clojure program we
must assume methods can return \clj{nil}.

In contrast, JVM invariants guarantee that  constructors cannot return \java{null},
so we are safe to assume constructors are non-nullable.

By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
but can be configured globally for particular positions if needed.

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that
returns the class of its argument or \clj{nil} if passed \clj{nil}.

We can use {\clj{defmethod}} to install a method to \clj{path}.
\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
Now the dispatch table maps
the \emph{dispatch value} \clj{String} to the function
\clj{(fn [x] x)}. 
We add another method
which maps
\clj{File} to the function
\clj{(fn [x] (.getPath x))}
in the dispatch table.
\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}

After installing both methods, the call 
$$
\clj{(path (new File "dir/a"))}
$$
dispatches to the second method we installed because
$$
\clj{(isa? (class "dir/a") String)}
$$
is true, and finally returns 
$$
\clj{((fn [x] (.getPath x)) "dir/a")}.
$$

The \clj{isa?} function first tries an equality check
on its arguments, then if that fails
and both arguments are classes a subclassing
check is returned.
\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? Keyword Object) ;=> true
\end{minted}

\clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
The reasoning is similar, except we only consider one branch (the current method) by
locally considering the current \emph{dispatch value} and reasoning about how it relates
to the \emph{dispatch function}.
For example, 
in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
and \clj{:const} methods.

>>>>

\subsection{Typed Clojure}

@@

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n]
  (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
\end{minted}

The function may take strings, but providing \clj{nil} (Clojure's name for Java's \java{null})
is a static type error --- \clj{nil} is not a string in Typed Clojure.

\begin{minted}{clj}
(greet nil)    ; Type Error:
               ;   Expected Str, given nil
               ; in: nil
\end{minted}

@@

All Clojure values are true except \clj{nil} and \clj{false}, so the
comma is only added when the argument is non-nil.

\paragraph{Flow analysis} Typed Clojure uses occurrence typing to
model type-based control flow.
In \clj{greetings}, a \clj{when} expression ensures \clj{repeat}
is never passed a \clj{nil} argument.

@@

Removing the \clj{when} expression is a static type error
--- \clj{repeat} cannot be passed \clj{nil}.

\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]
  (str "Hello, "
       (apply str (repeat i (str "hello, ")))
       ; Type Error:
       ;   Expected Num, given (U nil Num).
       ; in: i
       n "!"))
\end{minted}

@@

Clojure supports interoperability with Java, including the ability to
call constructors, invoke methods, and access fields.

\begin{minted}{clj}
  (.getParent (new File "a/b"))  ; :- (U nil Str)
  ;=> "a"
\end{minted}

If a specific constructor, method, or field cannot be found based on the
static types of its arguments, a type error is thrown.

\begin{minted}{clj}
  (fn [f] 
    (.getParent f)
    ; Type error:
    ;   Unresolved host interop: getParent
    )
\end{minted}

Function arguments default to \clj{Any}, the most permissive type. Ascribing
a parameter type helps Typed Clojure find a specific method.

%Calls to Java methods and fields have prefix notation
%like \clj{(.method target args*)} and \clj{(.field target)} respectively,
%with method and field names prefixed with a dot and methods taking some number of arguments.

%\begin{exmp}
%\inputminted[firstline=18,lastline=19]{clojure}{code/demo/src/demo/parent3.clj}
%\end{exmp}

The conditional guards from dereferencing \clj{nil}, and as before
removing it is a static type error.

\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)
  ; Type error:
  ;   Cannot call instance method 
  ;   java.io.File/getParent on type 
  ;   (U nil File).
  )
\end{minted}

Any Java method returning a reference can also return \java{null} ---
Typed Clojure rejects programs that assume otherwise.

\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)
  ; Type error:
  ;   Expected Str, given (U nil Str).
  ; in: (.getParent f)
  )
\end{minted}

%Typed Clojure and Java treat \java{null} differently.
%In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
%called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
%This means the Java static type \java{String} is equivalent to
%\clj{(U nil String)} in Typed Clojure.
%
%Reference types in Java are nullable, so to guarantee a method call does not
%leak \java{null} into a Typed Clojure program we
%must assume methods can return \clj{nil}.

In contrast, JVM invariants guarantee constructors return a non-null reference.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}

By default Typed Clojure conservatively assumes method and constructor arguments to be \emph{non-nullable},
but can be configured globally for particular non-target positions if needed.

\subsection{Multimethod}

A \emph{multimethod} is a kind of extensible function. It consists of
a \emph{dispatching function}, given with \clj{defmulti}, and a collection of \emph{methods} associated with
\emph{dispatch values}, each installed via \clj{defmethod}.

Recall the multimethod in figure~\ref{fig:ex1}.

\begin{minted}{clj}
(ann pname [(U File String) -> (U nil String)])
(defmulti pname class)
(defmethod pname String [s] (pname (new File s)))
(defmethod pname File [f] (.getName f))
\end{minted}

Its dispatching function is
\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
respectively.

When called, the dispatching function is applied to the provided arguments to
yield a \emph{dispatching value}.
The \clj{isa?} function is then used to compare the current dispatching value to each method's dispatch value.
An unambiguous match leads to the corresponding method being applied to the arguments,
giving the final result.

The following Typed Clojure program is semantically identical to figure~\ref{fig:ex1}.

\begin{minted}{clj}
(ann pname [(U Str File) -> (U nil Str)])
(defn pname [x]
  (let [dfn  class ; dispatching function
        dval (dfn x)] ; dispatching value
    (cond
      (and (isa? dval String)       ; if
           (not (isa? dval File))) 
      (pname (new File x))          ; then

      (and (not (isa? dval String)) ; else if
           (isa? dval File))       
      (.getName x)                  ; then
      :else (throw (Exception. "No match")))))
\end{minted}

The \clj{isa?} predicate 
is a subclassing check for classes, otherwise
an equality check.

\begin{minted}{clojure}
  (isa? :a :a) ;=> true
  (isa? :a :b) ;=> false
  (isa? String Object) ;=> true
  (isa? Object String) ;=> false
\end{minted}

%A multimethod in Clojure is a function with a \emph{dispatch
%function} and a \emph{dispatch table} of methods. Multimethods are created with {\clj{defmulti}}.
%\inputminted[firstline=5,lastline=6]{clojure}{code/demo/src/demo/rep.clj}
%The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
%and \emph{dispatch function} \clj{class}, a function that
%returns the class of its argument or \clj{nil} if passed \clj{nil}.
%
%We can use {\clj{defmethod}} to install a method to \clj{path}.
%\inputminted[firstline=7,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
%Now the dispatch table maps
%the \emph{dispatch value} \clj{String} to the function
%\clj{(fn [x] x)}. 
%We add another method
%which maps
%\clj{File} to the function
%\clj{(fn [x] (.getPath x))}
%in the dispatch table.
%\inputminted[firstline=8,lastline=8]{clojure}{code/demo/src/demo/rep.clj}
%
%After installing both methods, the call 
%$$
%\clj{(path (new File "dir/a"))}
%$$
%dispatches to the second method we installed because
%$$
%\clj{(isa? (class "dir/a") String)}
%$$
%is true, and finally returns 
%$$
%\clj{((fn [x] (.getPath x)) "dir/a")}.
%$$

The function \clj{inc-leaf} is like \egref{example:decmap} except the nodes are incremented.
The reasoning is similar, except we only consider one branch (the current method) by
locally considering the current \emph{dispatch value} and reasoning about how it relates
to the \emph{dispatch function}.
For example, 
in the \clj{:do} method we learn the \clj{:op} key is a \clj{:do}, which
narrows our argument type to the \clj{:do} Expr, and similarly for the \clj{:if}
and \clj{:const} methods.

@@
